---
title: "Assignment 5 Alt"
output: html_document
date: "2024-11-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(zoo)
```

## Load the data

```{r}
data <- readLines("https://www.nhc.noaa.gov/data/hurdat/hurdat2-nepac-1949-2016-041317.txt")

##parse_hurdat function taken from https://rdrr.io/cran/HURDAT/man/get_hurdat.html

hurdat <- parse_hurdat(data)
```

####Temperature data
```{r}
temp <- readLines("https://data.giss.nasa.gov/gistemp/graphs/graph_data/Global_Mean_Estimates_based_on_Land_and_Ocean_Data/graph.txt")

header_temp <- grep("^[A-Z]{2}[0-9]{6}", data)
event_temp <- setdiff(seq_along(data), header_lines)

header_temp <- data.frame(raw = temp[header_temp])
event_temp <- data.frame(raw = temp[event_temp])
event_temp <- event_temp %>% slice(-(1:4))

temp_df <- event_temp %>% separate(raw, into = c("Year", "No_Smoothing", "Lowess"), sep = "\\s+")

temp_df <- temp_df %>% filter(Year > 1948)
temp_df$Year <- as.numeric(temp_df$Year)

```{r}
hurdat <- hurdat %>% rename(Windspeed = Wind)

# hurdat <- hurdat %>%  
# mutate(
#     WindSpeed = as.numeric(WindSpeed),
#     Category = case_when(
#       WindSpeed < 64 ~ "Tropical Storm",
#       WindSpeed >= 64 & WindSpeed <= 82 ~ "Category 1",
#       WindSpeed >= 83 & WindSpeed <= 95 ~ "Category 2",
#       WindSpeed >= 96 & WindSpeed <= 113 ~ "Category 3",
#       WindSpeed >= 114 & WindSpeed <= 135 ~ "Category 4",
#       WindSpeed >= 136 ~ "Category 5"
#     )
#   )

hurdat <- hurdat %>% mutate(Year = format(DateTime, format = "%Y"))
```

## Merge Temp
```{r}
hurdat$Year <- as.numeric(hurdat$Year)
storm_temp <- hurdat %>% left_join(temp_df, by = "Year")
```

## Find Max Wind Speed by ID and separate hurricanes into Categories
```{r}
storm_temp_summary <- storm_temp %>% group_by(Key, Name, Year) %>% summarize(max_windspeed = max(Windspeed))

storm_temp_summary <- storm_temp_summary %>% left_join(temp_df, by = "Year")

storm_temp_summary <- storm_temp_summary %>%   mutate(
    max_windspeed = as.numeric(max_windspeed),
    Category = case_when(
      max_windspeed < 64 ~ "Tropical Storm",
      max_windspeed >= 64 & max_windspeed <= 82 ~ "Category 1",
      max_windspeed >= 83 & max_windspeed <= 95 ~ "Category 2",
      max_windspeed >= 96 & max_windspeed <= 113 ~ "Category 3",
      max_windspeed >= 114 & max_windspeed <= 135 ~ "Category 4",
      max_windspeed >= 136 ~ "Category 5"
    )
  )
```

##Find Frequency of Storms by Year
```{r}
storm_temp_freq <- storm_temp_summary %>%
  group_by(Year, Category) %>%
  summarize(freq = n(), .groups = 'drop')
storm_temp_freq <- storm_temp_freq %>% left_join(temp_df, by = "Year")

storm_temp_freq_overall <- storm_temp_summary %>%
  group_by(Year) %>%
  summarize(freq = n_distinct(Key))
```
## Average max windspeed per year
```{r}
windspeed_summary <- storm_temp_summary %>% group_by(Year) %>% summarise(avg_windspeed = mean(max_windspeed))
```


##Graph for Temp Changes

```{r}
ggplot(storm_temp_summary, aes(x = Year)) + 
  geom_line(aes(y = Lowess), linetype = "dashed", color = "#000000", size = 1) +  # Adjust size here
  labs(title = "Changes in Global Temperature and Wind Speed Over Time", 
       x = "Year",
       y = "Global Temperature Anomoly (째C)") + 
  theme_classic()
```


```{r}
storm_temp_summary$Lowess <- as.numeric(storm_temp_summary$Lowess)
scale_factor <- max(storm_temp_summary$Lowess, na.rm = TRUE) / max(storm_temp_summary$max_windspeed, na.rm = TRUE)

ggplot(storm_temp_summary, aes(x = Year)) + 
  geom_line(aes(y = Lowess), linetype = "solid", color = "blue") + 
  geom_point(aes(y = Lowess), color = "blue") +
  geom_line(aes(y = max_windspeed * scale_factor), linetype = "solid", color = "red") + 
  scale_y_continuous(
    name = "Global Temperature (째C)",
    sec.axis = sec_axis(~ . / scale_factor, name = "Average Wind Speed (scaled)")
  ) +
  labs(title = "Changes in Global Temperature and Wind Speed Over Time", 
       x = "Year") + 
  theme_classic() +
  theme(
    axis.title.y.left = element_text(color = "blue"),
    axis.title.y.right = element_text(color = "red"))


ggplot(windspeed_summary, aes(x = Year, y = avg_windspeed)) +
  geom_line(color = "red", alpha =.5) +
  geom_smooth(method = "loess", color = "blue") +
  labs(title = "Average Hurricane Windspeed Over Time", x = "Year", y = "Average Windspeed") +
  theme_classic()

```

```{r}
storm_temp_freq_overall <- storm_temp_freq_overall %>% left_join(temp_summary, by = "Year")


windspeed_summary <- windspeed_summary %>% left_join(temp_df, by = "Year")
windspeed_summary$Lowess <- as.numeric(windspeed_summary$Lowess)
windspeed_summary$No_Smoothing <- as.numeric(windspeed_summary$No_Smoothing)
cor.test(windspeed_summary$avg_windspeed, windspeed_summary$Lowess, method = "spearman")
cor.test(storm_temp_freq_overall$freq, storm_temp_freq_overall$avg_temp, method = "spearman")

ggplot(windspeed_summary, aes(x = avg_windspeed, y = Lowess)) +
  geom_point() +
  geom_smooth(method = "loess", color = "blue") +  
  labs(
    title = "Relationship Between Average Wind Speed and Global Temperature",
    x = "Average Wind Speed",
    y = "Anomoly Temperature (째C)"
  ) +
  theme_classic()

ggplot(storm_temp_freq_overall, aes(x = freq, y = avg_temp)) +
  geom_point() +
  geom_smooth(method = "loess", color = "blue") +  
  labs(
    title = "Relationship Between Storm Frequency and Global Temperature",
    x = "Strom Frequency",
    y = "Anomoly Temperature (째C)"
  ) +
  theme_classic()
```

## Plotting the Frequency of Hurricane Types over the Years
```{r, fig.width=12, fig.height=8}
storm_temp_freq <- storm_temp_freq %>% filter(!is.na(Category))

storm_temp_freq$Lowess <- as.numeric(storm_temp_freq$Lowess)

freq_range <- range(storm_temp_freq$freq, na.rm = TRUE)
lowess_range <- range(storm_temp_freq$Lowess, na.rm = TRUE)

ggplot(storm_temp_freq, aes(Year)) + 
  geom_smooth(aes(y = freq, color = Category, fill = Category), method = "loess", se = FALSE, size = 1) +
  geom_line(aes(y = scales::rescale(Lowess, to = freq_range), group = 1), color = "black", size = 0.8, linetype = "dashed") +
  facet_wrap(~ Category, nrow = 2) +
  scale_y_continuous(name = "Storm Frequency (Smoothed)") +  # Remove secondary axis here
  labs(x = "Year", title = "Storm Frequency and Global Temperature Changes by Storm Category") +
  theme_classic() +
  theme(legend.position = "bottom")
```
```{r}
storm_temp_mov_avg<- storm_temp_freq %>%
  group_by(Category) %>%
  arrange(Year) %>%
  mutate(freq_ma = rollmean(freq, k = 5, fill = NA, align = "right"))

freq_range <- range(storm_temp_mov_avg$freq, na.rm = TRUE)
lowess_range <- range(storm_temp_mov_avg$Lowess, na.rm = TRUE)

# Plot with pre-calculated ranges
ggplot(storm_temp_mov_avg, aes(Year)) + 
  geom_col(aes(y = freq, color = Category, fill = Category), position = position_dodge(), alpha = 0.6) +
  geom_line(aes(y = freq_ma, group = Category, color = Category), size = 1) +
  geom_line(aes(y = scales::rescale(Lowess, to = freq_range), group = 1), color = "black", size = 0.8, linetype = "dashed") +
  facet_wrap(~ Category, nrow = 2) +
  scale_y_continuous(
    sec.axis = sec_axis(
      ~ scales::rescale(., from = freq_range, to = lowess_range), 
      name = "Temperature"
    )
  ) +
  labs(x = "Year", y = "Frequency", title = "Hurricane Frequency (5-Year Moving Average) and Temperature Trends by Category") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

```{r}
ggplot(storm_temp_summary, aes(x = Year)) + geom_histogram()

ggplot(storm_temp_freq_overall, aes(x = Year, y = freq)) + geom_line(color = "red", alpha =.5) +
  geom_smooth(method = "loess", color = "blue") +
  labs(title = "Hurricane Frequency Over Time", x = "Year", y = "Frequency") +
  theme_classic()
```

